Quantitative Estimates of Minimum Core Masses for Giant Planet Formation {\bf Paper I was quantitative.}

### 2013-11-20 tex/draft_paper2/draft_paper2.tex
section 4 comments

\usepackage{graphicx} 
\usepackage{graphics}                       
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{subeqnarray}
%\bibliographystyle{apj}

@@

\title{Quantitative Estimates of Minimum Core Masses for Giant Planet Formation}
\author{Ana-Maria A. Piso}
\affil{Harvard-Smithsonian Center for Astrophysics}
\author{Andrew N. Youdin}
\affil{JILA, University of Colorado at Boulder}
\author{Ruth A. Murray-Clay}
\affil{Harvard-Smithsonian Center for Astrophysics}

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/delad_S_mixt.pdf}
%%\vspace{-0.5in}
\caption{Contour plot of the adiabatic gradient $\delad$ for a hydrogen-helium mixture as a function of gas temperature and pressure. The upper-right rectangle encloses the region described by the original \citet{saumon95} EOS tables, while the rest of the plot is our extension. The black curves represent constant entropy adiabats, with the labels $\log_{10}(S)$ [erg K$^{-1}$ g$^{-1}$]. The regions in which the EOS is either invalid or not computed are masked in white. Our extension is only valid for $T \lesssim 2000$ K, since it does not take into account hydrogen dissociation. We choose $T=1500$ K as a conservative temperature cutoff. \citet{saumon95} do not compute the EOS at very high pressures, since hydrogen is solid or may form a Coulomb lattice in this regime, and thus their EOS treatment is no longer valid. While the boundaries of the region in which the free-energy EOS treatment fails can be determined from fundamental thermodynamic constraints, such calculations are not the object of this work. Instead, we choose as boundary a constant entropy curve ($\log(S)=8.4$) above the region in which the \citet{saumon95} model fails. The expressions derived in \App{EOStables} are sufficient to give good results for the colored regions of the extended map, which fully cover the temperature and pressures ranges required by our models.}
\label{fig:deladmap}
\end{figure}
%but not the white ones. We note, however, that the temperature and pressure ranges required by our models are fully covered by the colored regions, where our expressions are valid

@@

\section{Equation of State Effects on Atmosphere Evolution}
\label{EOSeffects}

Variations in the adiabatic gradient $\delad$ due to partial dissociation and quantum rotational effects (see \S\ref{deladtable}) have two competing effects on the evolution of a core's atmosphere. They yield: (1) a lower envelope luminosity, and (2) a larger amount of radiated energy per unit of accreted mass, when compared to an ideal gas. Both effects combined slow down accretion and increase the growth time of the atmosphere and thus the critical core mass.

In order to explain how quantum rotational states at low temperatures affect atmospheric growth, we perform the following thought experiment: we imagine that the EOS deviates from a polytrope only in the upper atmosphere where temperatures are low, and that the gas is ideal and polytropic deep in the envelope. Conversely, we study the effects of dissociation at high temperatures by assuming that the EOS is realistic at the bottom of the atmosphere where temperatures are high, but that the gas is polytropic in the outer regions. Figure \ref{fig:tplotall} shows the time evolution of atmospheres forming at 10 AU around cores of mass $M\co=10 M_{\oplus}$ and described by various equations of state, as follows:

\begin{itemize}
\item Dashed curve (1) --- ideal gas polytrope with $\delad=0.3$.
\item Dotted curve (2) --- ideal gas polytrope with $\delad=0.3$ for $T>500$ K and realistic EOS for $T<500$ K.
\item Dash-dotted curve (3) --- ideal gas polytrope with $\delad=0.3$ for $T<500$ and realistic EOS for $T>500$ K.
\item Solid curve (4) --- fully realistic EOS.
\end{itemize}
We choose $T=500$ K as the cutoff temperature because the hydrogen-helium mixture behaves like an ideal gas, with $\delad=0.3$, in this temperature regime (see Figure \ref{fig:deladmap}). As such, the time evolution difference between curves (1) and (2) shows the quantum rotational effects on the atmosphere evolution, while the difference between curves (1) and (3) accounts for hydrogen dissociation. We see that both dissociation and rotational states have a comparable effect on the atmosphere growth and result in slower cooling. The growth time is dependent on both the total energy that must be released (i.e., the rate at which energy is released) and on the luminosity of the atmosphere. We explore the relative influence of these two factors separately.

@@

\begin{figure*}[tb]
\centering
\includegraphics[width=\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/all_plot.pdf}
%%\vspace{-0.5in}
\caption{We explore atmosphere growth around a core of $M\co=10 M_{\oplus}$ forming at 10 AU in our fiducial disk, for the EOS choices in Figure \ref{fig:tplotall}. Upper-left panel:  Instantaneous temperature profile as a function of the radial coordinate when the total mass (core + atmosphere) is $12 M_{\oplus}$. The location of the RCB is marked. Quantum rotational states at low temperatures in the outer region of the atmosphere result in a deeper RCB. Upper-right panel:  Luminosity evolution as a function of total mass. Quantum rotational effects at low temperatures near the top of the atmosphere result in a lower luminosity for the realistic EOS (see text). Lower-left panel: Instantaneous energy profiles as a function on radius for a total mass of $12 M_{\oplus}$. Hydrogen dissociation deep in the atmosphere causes the bulk of the energy to be concentrated at the bottom of the envelope. This increases the amount of energy per unit mass that needs to be radiated away ($|dE/dM|$) resulting in a longer crossover time for the realistic EOS when compared to the polytrope. Lower-right panel:  $|dE/dM|$ evolution as a function of total mass. Hydrogen dissociation deep in the atmosphere results in the bulk of the energy being concentrated at the bottom of the envelope, which increases the energy per unit mass that has to be radiated away when compared to an ideal gas polytrope. This results in slower growth.}
\label{fig:all_plot}
\end{figure*}

The upper-right panel of Figure \ref{fig:all_plot} shows the luminosity evolution with mass for atmospheres (1)-(4), for an example core with $M\co=10 M_{\oplus}$ forming at 10 AU. We see that the lower luminosity of the fully realistic EOS (4) when compared to the polytrope (1) is due to the realistic EOS in the outer atmosphere (2). This behavior is directly correlated to the depth of the radiative region of the envelope, which is shown in the upper-left panel of Figure \ref{fig:all_plot} for a total planet mass of $12 M_{\oplus}$. We see that the realistic EOS for low temperatures (2) results in a deeper location of the RCB when compared to the ideal gas polytrope (1). This deeper RCB translates into a larger number of steps that photons need to diffuse to escape from the RCB, and therefore a lower luminosity (cf. Equation \ref{eq:delrad}).

@@

Dissociation deep in the planet's atmosphere increases the amount of energy per unit mass that needs to be radiated by the envelope, which further slows down growth. Figure \ref{fig:all_plot}, lower-left plot, shows the total energy (internal and gravitational) profile as a function of the radial coordinate, for the same example planet. The bulk of the energy is concentrated deep in the atmosphere for the realistic EOS at high temperatures (3) and the fully realistic EOS curve (4), which shows that hydrogen dissociation in the inner atmosphere dictates the energy behavior of the envelope. In contrast, energy is concentrated towards the outer boundary for the $\delad=0.3$ polytrope (1). Qualitatively, this can be explained through a simple analytic argument: the density profile in an adiabatic, non-self-gravitating atmosphere composed of an ideal gas scales as $\rho(r) \sim r^{1/\delad-1}$ (see Paper I). Since the total energy scales as $E(r) \sim \rho(r) r^2 \sim r^{1/\delad+1}$, adiabats with lower $\delad$ have their energy more tightly packed towards the interior of the envelope, which is the case during dissociation when $\delad$ decreases significantly (see Section \S\ref{deladtable} and Figure \ref{fig:deladmap}).  It takes more energy to bring gas deep into the atmosphere for an envelope that has the bulk of its energy concentrated towards the bottom,  which increases the amount of energy per unit mass that needs to be radiated, i.e. $|dE/dM|$. The cooling time is therefore larger for the realistic EOS at high temperatures (3) when compared to the ideal gas polytrope (1). This increase in $|dE/dM|$ is shown in the bottom-right panel of Figure \ref{fig:all_plot}.

We have seen that quantum rotational states transitions due to the low temperatures at the top of the atmosphere increase the thickness of the radiative zone, and decrease the luminosity of the envelope, while dissociation at high temperatures deep in the atmosphere increases the amount of energy needed to increase the atmosphere mass by a fixed amount. Overall, both effects result in a longer time for the atmosphere to evolve.

@@

\acknowledgements{We thank Sean Andrews for providing us with the \citet{dalessio01} opacity tables. We also thank Eugene Chiang and Steven Cranmer for useful conversations. ANY thanks Phil Arras for his help with entropy units conventions in the equation of state tables.}

>>>>

\usepackage{graphicx} 
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{subeqnarray}
\usepackage{ulem}
%\bibliographystyle{apj}

@@

\title{Quantitative Estimates of Minimum Core Masses for Giant Planet Formation
{\bf Paper I was quantitative.}}
\author{Ana-Maria A. Piso}
\affil{Harvard-Smithsonian Center for Astrophysics}
\author{Andrew N. Youdin}
\affil{JILA, University of Colorado at Boulder}
\author{Ruth A. Murray-Clay}
\affil{Harvard-Smithsonian Center for Astrophysics}

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/delad_S_mixt.pdf}
%%\vspace{-0.5in}
\caption{Contour plot of the adiabatic gradient $\delad$ for a hydrogen-helium mixture as a function of gas temperature and pressure. The upper-right rectangle encloses the region described by the original \citet{saumon95} EOS tables, while the rest of the plot is our extension. The black curves represent constant entropy adiabats, with the labels $\log_{10}(S)$ [erg K$^{-1}$ g$^{-1}$]. The regions in which the EOS is either invalid or not computed are masked in white. Our extension is only valid for $T \lesssim 2000$ K, since it does not take into account hydrogen dissociation. We choose $T=1500$ K as a conservative temperature cutoff. \citet{saumon95} do not compute the EOS at very high pressures, since hydrogen is solid or may form a Coulomb lattice in this regime, and thus their EOS treatment is no longer valid. While the boundaries of the region in which the free-energy EOS treatment fails can be determined from fundamental thermodynamic constraints, such calculations are not the object of this work. Instead, we choose as boundary a constant entropy curve ($\log(S)=8.4$) above the region in which the \citet{saumon95} model fails. The expressions derived in \App{EOStables} are sufficient to give good results for the colored regions of the extended map, which fully cover the temperature and pressures ranges required by our models. {\bf This caption is extremely long.  I would find a way to defer much of this to text.}}
\label{fig:deladmap}
\end{figure}
%but not the white ones. We note, however, that the temperature and pressure ranges required by our models are fully covered by the colored regions, where our expressions are valid

@@

\section{Equation of State Effects on Atmosphere Evolution \\
{\bf long section title.  Just "Role of EOS"?}}
\label{EOSeffects}

Variations in the adiabatic gradient $\delad$ due to partial dissociation and quantum {\bf (`Quantum' is not a useful descriptor (IMO).  All EOS effects are quantum, noting molecular or H$_2$ rotation would be more descriptive.)} rotational effects (see \S\ref{deladtable}) have two competing {\bf (Not `competing,' both contribute and can/do go in same direction)} effects on \sout{the} {\bf atmospheric} evolution \sout{of a core's atmosphere}.  {\bf Putting the equation $\dot{M} = -L/(dE/dM)$ early in the explanation would help.}
They yield: (1) a lower envelope luminosity, and (2) a larger amount of radiated energy per unit of accreted mass, when compared to an ideal gas.  Both effects {\bf A bit confusing if you mean EOS effects or resulting evolution effects.  Maybe avoid reusing `effects`.)} combined slow down accretion and increase the growth time of the atmosphere and thus the critical core mass.

{\bf In what follows it may be clearer to start with the motivation: isolating the two EOS effects, i.e. dissociation and rotation.  Then introduce the two patched EOSs as how we do this.  Currently it's less clear and more drawn out where this is going.}

In order to explain how quantum rotational states at low temperatures affect atmospheric growth, we perform the following thought experiment {\bf (I don't see how this is a thought experiment, or why we have to imagine.  I would just describe the EOSs)}: we imagine that the EOS deviates from a polytrope only in the upper atmosphere where temperatures are low, and that the gas is ideal and polytropic deep in the envelope. Conversely, we study the effects of dissociation at high temperatures by assuming that the EOS is realistic at the bottom of the atmosphere where temperatures are high, but that the gas is polytropic in the outer regions. {{\bf Finish the description of the EOSs first.  Then describe and interpret Fig. 2 starting in a new paragraph.  Starting the fig description at the end of a methods discussion lost me.)}  Figure \ref{fig:tplotall} shows the time evolution of atmospheres forming at 10 AU around cores of mass $M\co=10 M_{\oplus}$ and described by various equations of state, as follows:

{\bf I would make this list numbered and omit the reference to the curve styles, which is clear in figures.}
\begin{itemize}
\item Dashed curve (1) --- ideal gas polytrope with $\delad=0.3$.
\item Dotted curve (2) --- ideal gas polytrope with $\delad=0.3$ for $T>500$ K and realistic EOS for $T<500$ K.
\item Dash-dotted curve (3) --- ideal gas polytrope with $\delad=0.3$ for $T<500$ and realistic EOS for $T>500$ K.
\item Solid curve (4) --- fully realistic EOS. ({\bf Here and elsewhere, I don't like ``fully" realistic, as it implies a perfect EOS.  I would just say realistic at all $T$.})
\end{itemize}
We choose $T=500$ K as the cutoff temperature because the hydrogen-helium mixture behaves like an ideal gas, with $\delad=0.3$, in this temperature regime (see Figure \ref{fig:deladmap}). {\bf (I find the references to curve numbers distracting in general.  I would try to do less of this and focus on describing the physical effects.  The reader still needs to look at the figure, and there the curves are clearly labelled.)} \sout{As such}, the time evolution difference between curves (1) and (2) shows the quantum rotational effects on the atmosphere evolution, while the difference between curves (1) and (3) accounts for hydrogen dissociation. We see that both dissociation and rotational states have a comparable effect {\bf (due to log scale, I would not call the effects comparable, just b/c they go in same direction)} on the atmosphere growth and result in slower cooling. The growth time is dependent on both the total energy that must be released (i.e., the rate at which energy is released) and on the luminosity of the atmosphere. {\bf `Rate of energy release' and luminosity are the same thing.  You may be calling $dE/dM$ a `rate' which is confusing.  Again, I think if you introduce and physically explain $\dot{M} = -L/(dE/dM)$ at the outset, then the resulting explanations can more compactly and clearly refer to the terms that matter.}  We explore the \sout{relative} influence of these two factors separately.  {\bf Haven't you already started exploring these factors separately?  Why say this again? Also perhaps worth a note (if not an explanation) that the combined effect is significantly greater than either individually (again noting log scale).}

@@

\begin{figure*}[tb]
\centering
\includegraphics[width=\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/all_plot.pdf}
%%\vspace{-0.5in}
\caption{{\bf (Left and right panels could be different figs.  The left panel could benefit from showing $P$ or $\rho$.  Also $E_{\rm tot}$ is not yet well defined; the local quantity $e \rho r^3$ might be more instructive.  A greater range of $E$ would show what the polytrope is doing better.  The right panel could benefit from adding $\dot{M}$.  We call $E_{\rm tot} \rightarrow E$ elsewhere, so should be consistent.  Why don't x-axes line up in right panel, are masses different?  What is the discontinuity in $dE/dM$ for the polytrope around 14 $M_\oplus$?  (A kink would be course resolution, but a discontinuity is harder to understand.)  I don't think you need an abs.\ val. around $-dE/dM$.)} We explore atmosphere growth around a core of $M\co=10 M_{\oplus}$ forming at 10 AU in our fiducial disk, for the EOS choices in Figure \ref{fig:tplotall}. Upper-left panel:  Instantaneous temperature profile as a function of the radial coordinate when the total mass (core + atmosphere) is $12 M_{\oplus}$. The location of the RCB is marked. Quantum rotational states at low temperatures in the outer region of the atmosphere result in a deeper RCB. Upper-right panel:  Luminosity evolution as a function of total mass. Quantum rotational effects at low temperatures near the top of the atmosphere result in a lower luminosity for the realistic EOS (see text). Lower-left panel: Instantaneous energy profiles as a function on radius for a total mass of $12 M_{\oplus}$. Hydrogen dissociation deep in the atmosphere causes the bulk of the energy to be concentrated at the bottom of the envelope. This increases the amount of energy per unit mass that needs to be radiated away ($|dE/dM|$) resulting in a longer crossover time for the realistic EOS when compared to the polytrope. Lower-right panel:  $|dE/dM|$ evolution as a function of total mass. Hydrogen dissociation deep in the atmosphere results in the bulk of the energy being concentrated at the bottom of the envelope, which increases the energy per unit mass that has to be radiated away when compared to an ideal gas polytrope. This results in slower growth.}
\label{fig:all_plot}
\end{figure*}

The upper-right panel of Figure \ref{fig:all_plot} shows the luminosity evolution with mass for \sout{atmospheres (1)-(4)} {\bf our model EOSs}, for a\sout{n example core with} $M\co=10 M_{\oplus}$ {\bf core} \sout{forming} {\bf (as core isn't forming, this phrase is misleading in addition to being too lengthy)} at 10 AU. We see that the lower luminosity of the fully realistic EOS (4) when compared to the polytrope (1) is due to the realistic EOS in the outer atmosphere (2). This behavior is directly correlated to the depth of the radiative region of the envelope, which is shown in the upper-left panel of Figure \ref{fig:all_plot} for a total planet mass of $12 M_{\oplus}$. We see that the realistic EOS for low temperatures (2) results in a deeper location of the RCB when compared to the ideal gas polytrope (1). {\bf This is an accurate observation.  
The deeper question is why does the EOS change result in a deeper RCB?  This should be possible to understand, right?}  This deeper RCB translates into a larger number of steps that photons need to diffuse to escape from the RCB, and therefore a lower luminosity (cf. Equation \ref{eq:delrad}).  {\bf Better to just say it's an optical depth effect than reexplaining how radiative diffusion works.  Also it's the pressure depth (not the radius) that correlates more closely with optical depth.  A plot of $P(r)$ might thus help as noted in fig caption.}

@@

Dissociation deep in the planet's atmosphere increases the amount of energy per unit mass that needs to be radiated by the envelope, which further slows down growth. 
{\bf Why is this true?  Dissociation absorbs energy which by itself decreases the amount of energy that needs to be radiated.}  Figure \ref{fig:all_plot}, lower-left plot, shows the total energy (internal and gravitational) profile as a function of the radial coordinate, \sout{for the same example planet}. {\bf Two choices: (1) define this cumulative energy more clearly and give it a different symbol than $E_{\rm tot}$ that's already used, maybe $E(r)$, or (2) plot the local (noncumulative) quantity $e \rho r^3$, which should peak in the interesting places.} The bulk of the energy is concentrated deep in the atmosphere for the realistic EOS at high temperatures (3) and the fully realistic EOS curve (4), which shows that hydrogen dissociation in the inner atmosphere dictates the energy behavior of the envelope. In contrast, energy is concentrated towards the outer boundary for the $\delad=0.3$ polytrope (1). 
{\bf The key question here is how does the slope of the energy affect the quantity that matters, $|dE/dM|$?  This is not clear yet.  I think the shift to lower temperatures, seen in the $T(r)$ plot is helpful, but perhaps the EOS plots should also include an internal energy map.}
Qualitatively, this can be explained through a simple analytic argument: the density profile in an adiabatic, non-self-gravitating atmosphere composed of an ideal gas scales as $\rho(r) \sim r^{1/\delad-1}$ (see Paper I) {\bf (The sign is wrong here and the error propagates to the energy.  Should be $\propto$ not $\sim$.)}. Since the total energy scales as $E(r) \sim \rho(r) r^2 \sim r^{1/\delad+1}$, {\bf (This is a new/contradictory use of $E$, what is given here is $dE/d\ln(r)$.  Scaling is wrong as noted above.  Again $\propto$ not $\sim$.)} adiabats with lower $\delad$ have their energy more tightly packed towards the interior of the envelope, which is the case during dissociation when $\delad$ decreases significantly (see Section \S\ref{deladtable} and Figure \ref{fig:deladmap}).  It takes more energy to bring gas deep into the atmosphere for an envelope that has the bulk of its energy concentrated towards the bottom,  {\bf (This argument isn't clear, new gas isn't brought deep into the atmosphere, it is added at the surface.)} which increases the amount of energy per unit mass that needs to be radiated, i.e. $|dE/dM|$. The cooling time is therefore larger {\bf with dissociation} \sout{for the realistic EOS at high temperatures} (3) when compared to the ideal gas polytrope (1). This increase in $|dE/dM|$ is shown in the bottom-right panel of Figure \ref{fig:all_plot}.

We have seen that quantum rotational states transitions due to the low temperatures at the top of the atmosphere {\bf (already too long and convoluted)} increase the thickness of the radiative zone, and decrease the luminosity of the envelope, while dissociation at high temperatures deep in the atmosphere increases the amount of energy needed to increase the atmosphere mass by a fixed amount. {\bf Way too long a sentence!} Overall, both effects result in a longer time for the atmosphere to evolve.

@@

\acknowledgements{We thank Sean Andrews for providing us with the \citet{dalessio01} opacity tables. We also thank Eugene Chiang and Steven Cranmer for useful conversations. ANY thanks Phil Arras for a methodical explanation of entropy.}
